---
title: Final lab survival analysis
author: "AS Huffmyer"
date: '2026'
output:
  html_document:
    code_folding: hide
    toc: yes
    toc_depth: 6
    toc_float: yes
  pdf_document:
    keep_tex: yes
editor_options: 
  chunk_output_type: console
---

# Set up 

Set up workspace, set options, and load required packages.    
```{r}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Load libraries. 
```{r}
library(tidyverse)
library(ggplot2)
library(survival)
library(readxl)
library(ggsurvfit)
library(gtsummary)
library(cardx)
library(cowplot)
library(lme4)
library(lmerTest)
library(car)
library(emmeans)
```

# Read in data 

Read in data. 
```{r}
data<-read_csv("data/final-lab-testing/final_lab_survival.csv")%>%
  mutate(time=as.numeric(time))
```

Read in metadata. 
```{r}
metadata<-read_csv("data/final-lab-testing/bag_metadata.csv")%>%select(!c(field_tags,live, dead))
```

Merge in metadata. 
```{r}
data<-left_join(data, metadata)
head(data)
```

"treatment" here refers to the original hardening treatment. All oysters were run in the survival assay at 34-36°C.  

# Plot temperatures 

Plot temperatures for each round 

```{r}
temps<-read_csv("data/final-lab-testing/temperature.csv")
```

Set theme. 
```{r}
my_theme<-theme_classic()+
  theme(
    axis.title.y = element_text(size = 14, color = "black"),
    axis.text.y = element_text(size=12, color = "black"),
    axis.text.x = element_text(size=12, color = "black"),
    axis.title.x =element_text(size = 14, color = "black"),
    legend.position = "none")
```

Plot faceted by round. 

```{r}
plot<-temps%>%
  
  ggplot(aes(x=time, y=temp.C, color=round))+
  facet_grid(~round)+
  geom_point()+
  stat_summary(fun = mean, geom = "line", linewidth = 1)+
  my_theme+
  theme(legend.position="right");plot

ggsave(filename="figures/final-lab-survival/temperature_rounds.png", width=10, height=5)
```

Round 2 had a temperature control issue - we will likely need to exclude this round. Round 1 was in a waterbath so temperatures started warm. Rounds 3-5 were in an incubator and therefore had a warm up time once the trial started.  

# Generate survival curves

```{r}
s1 <- survival::survfit(Surv(time, status) ~ treatment+family, data = data)
str(s1)
```

Plot the survival function
```{r}
ggsurvfit::survfit2(Surv(time, status) ~ treatment+family, data = data) %>% 
  ggsurvfit::ggsurvfit() +
  labs(
    x = "Hours",
    y = "Survival probability"
  )
```

Estimate survival at 48 hours. 
```{r}
summary(survival::survfit(Surv(time, status) ~ treatment+family, data = data), times = 48)
```

Use a log rank model to determine statistical differences in curves. 
```{r}
survival::survdiff(Surv(time, status) ~ treatment+family, data = data)
```

Analyze again with a Cox proportional hazards model. 
```{r}
survival::coxph(Surv(time, status) ~ treatment*family+round, data = data)

survival::coxph(Surv(time, status) ~ treatment*family+round, data = data) %>% 
  tbl_regression(exp = TRUE) 
```

## Generate plots 

### Treatment 

Round 1
```{r}
plot1<-ggsurvfit::survfit2(Surv(time, status) ~ treatment, data = data%>%filter(round=="round1")) %>% 
  ggsurvfit::ggsurvfit() +
  labs(
    x = "Hours",
    y = "Survival probability",
    title=""
  )+
  ylim(0,1)+
  my_theme+
  #scale_linetype_manual(values=c("solid", "dashed"))+
  scale_colour_manual(values=c("darkgray", "orange"))+
  theme(legend.position="right");plot1
```

Round 2
```{r}
plot1<-ggsurvfit::survfit2(Surv(time, status) ~ treatment, data = data%>%filter(round=="round2")) %>% 
  ggsurvfit::ggsurvfit() +
  labs(
    x = "Hours",
    y = "Survival probability",
    title=""
  )+
  ylim(0,1)+
  my_theme+
  #scale_linetype_manual(values=c("solid", "dashed"))+
  scale_colour_manual(values=c("darkgray", "orange"))+
  theme(legend.position="right");plot1
```

Round 3
```{r}
plot1<-ggsurvfit::survfit2(Surv(time, status) ~ treatment, data = data%>%filter(round=="round3")) %>% 
  ggsurvfit::ggsurvfit() +
  labs(
    x = "Hours",
    y = "Survival probability",
    title=""
  )+
  ylim(0,1)+
  my_theme+
  #scale_linetype_manual(values=c("solid", "dashed"))+
  scale_colour_manual(values=c("darkgray", "orange"))+
  theme(legend.position="right");plot1
```

Round 4
```{r}
plot1<-ggsurvfit::survfit2(Surv(time, status) ~ treatment, data = data%>%filter(round=="round4")) %>% 
  ggsurvfit::ggsurvfit() +
  labs(
    x = "Hours",
    y = "Survival probability",
    title=""
  )+
  ylim(0,1)+
  my_theme+
  #scale_linetype_manual(values=c("solid", "dashed"))+
  scale_colour_manual(values=c("darkgray", "orange"))+
  theme(legend.position="right");plot1
```

Round 5
```{r}
plot1<-ggsurvfit::survfit2(Surv(time, status) ~ treatment, data = data%>%filter(round=="round5")) %>% 
  ggsurvfit::ggsurvfit() +
  labs(
    x = "Hours",
    y = "Survival probability",
    title=""
  )+
  ylim(0,1)+
  my_theme+
  #scale_linetype_manual(values=c("solid", "dashed"))+
  scale_colour_manual(values=c("darkgray", "orange"))+
  theme(legend.position="right");plot1
```

All rounds
```{r}
plot1<-ggsurvfit::survfit2(Surv(time, status) ~ treatment, data = data) %>% 
  ggsurvfit::ggsurvfit() +
  labs(
    x = "Hours",
    y = "Survival probability",
    title=""
  )+
  ylim(0,1)+
  my_theme+
  #scale_linetype_manual(values=c("solid", "dashed"))+
  scale_colour_manual(values=c("darkgray", "orange"))+
  theme(legend.position="right");plot1
```

### Family 

Round 1
```{r}
plot2<-ggsurvfit::survfit2(Surv(time, status) ~ family, data = data%>%filter(round=="round1")) %>% 
  ggsurvfit::ggsurvfit() +
  labs(
    x = "Hours",
    y = "Survival probability",
    title=""
  )+
  ylim(0,1)+
  my_theme+
  theme(legend.position="right");plot2
```

Round 2
```{r}
plot2<-ggsurvfit::survfit2(Surv(time, status) ~ family, data = data%>%filter(round=="round2")) %>% 
  ggsurvfit::ggsurvfit() +
  labs(
    x = "Hours",
    y = "Survival probability",
    title=""
  )+
  ylim(0,1)+
  my_theme+
  theme(legend.position="right");plot2
```

Round 3
```{r}
plot2<-ggsurvfit::survfit2(Surv(time, status) ~ family, data = data%>%filter(round=="round3")) %>% 
  ggsurvfit::ggsurvfit() +
  labs(
    x = "Hours",
    y = "Survival probability",
    title=""
  )+
  ylim(0,1)+
  my_theme+
  theme(legend.position="right");plot2
```

Round 4
```{r}
plot2<-ggsurvfit::survfit2(Surv(time, status) ~ family, data = data%>%filter(round=="round4")) %>% 
  ggsurvfit::ggsurvfit() +
  labs(
    x = "Hours",
    y = "Survival probability",
    title=""
  )+
  ylim(0,1)+
  my_theme+
  theme(legend.position="right");plot2
```

Round 5
```{r}
plot2<-ggsurvfit::survfit2(Surv(time, status) ~ family, data = data%>%filter(round=="round5")) %>% 
  ggsurvfit::ggsurvfit() +
  labs(
    x = "Hours",
    y = "Survival probability",
    title=""
  )+
  ylim(0,1)+
  my_theme+
  theme(legend.position="right");plot2
```

All rounds
```{r}
plot2<-ggsurvfit::survfit2(Surv(time, status) ~ family, data = data) %>% 
  ggsurvfit::ggsurvfit() +
  labs(
    x = "Hours",
    y = "Survival probability",
    title=""
  )+
  ylim(0,1)+
  my_theme+
  theme(legend.position="right");plot2
```

### Treatment group x family 

Round1
```{r}
plot3<-ggsurvfit::survfit2(Surv(time, status) ~ family+treatment, data = data%>%filter(round=="round1")) %>% 
  ggsurvfit::ggsurvfit() +
  labs(
    x = "Hours",
    y = "Survival probability",
    title=""
  )+
  ylim(0,1)+
  my_theme+
  theme(legend.position="right");plot3
```

Round2
```{r}
plot3<-ggsurvfit::survfit2(Surv(time, status) ~ family+treatment, data = data%>%filter(round=="round2")) %>% 
  ggsurvfit::ggsurvfit() +
  labs(
    x = "Hours",
    y = "Survival probability",
    title=""
  )+
  ylim(0,1)+
  my_theme+
  theme(legend.position="right");plot3
```

Round3
```{r}
plot3<-ggsurvfit::survfit2(Surv(time, status) ~ family+treatment, data = data%>%filter(round=="round3")) %>% 
  ggsurvfit::ggsurvfit() +
  labs(
    x = "Hours",
    y = "Survival probability",
    title=""
  )+
  ylim(0,1)+
  my_theme+
  theme(legend.position="right");plot3
```

Round4
```{r}
plot3<-ggsurvfit::survfit2(Surv(time, status) ~ family+treatment, data = data%>%filter(round=="round4")) %>% 
  ggsurvfit::ggsurvfit() +
  labs(
    x = "Hours",
    y = "Survival probability",
    title=""
  )+
  ylim(0,1)+
  my_theme+
  theme(legend.position="right");plot3
```

Round5
```{r}
plot3<-ggsurvfit::survfit2(Surv(time, status) ~ family+treatment, data = data%>%filter(round=="round5")) %>% 
  ggsurvfit::ggsurvfit() +
  labs(
    x = "Hours",
    y = "Survival probability",
    title=""
  )+
  ylim(0,1)+
  my_theme+
  theme(legend.position="right");plot3
```

All rounds
```{r}
plot3<-ggsurvfit::survfit2(Surv(time, status) ~ family+treatment, data = data) %>% 
  ggsurvfit::ggsurvfit() +
  labs(
    x = "Hours",
    y = "Survival probability",
    title=""
  )+
  ylim(0,1)+
  my_theme+
  theme(legend.position="right");plot3

```

Plot family 164 specifically. 

```{r}
plot<-ggsurvfit::survfit2(Surv(time, status) ~ family+treatment, data = data%>%filter(family=="YC24-164")) %>% 
  ggsurvfit::ggsurvfit() +
  labs(
    x = "Hours",
    y = "Survival probability",
    title=""
  )+
  ylim(0,1)+
  my_theme+
  theme(legend.position="right");plot
```

# Run binomial logistic models 

Run a binomial logistic model to account for variation between rounds. Include round and oyster size (length in mm) as random effects. 

0 = alive
1 = mortality 

```{r}
surv_model<-glmer(status ~ time * family * treatment + (1|round), family = binomial(link = "logit"), data=data)

summary(surv_model)
Anova(surv_model)

emm<-emmeans(surv_model, ~treatment|family)
pairs(emm)

emm<-emmeans(surv_model, ~family)
pairs(emm)
```

Generate predictive mortality plot. 

```{r}
# Generate predicted probabilities
data$predicted_mortality <- predict(surv_model, type = "response")

# Plot
plot2<-ggplot(data, aes(x = time, y = predicted_mortality, color = treatment, fill = treatment, group=treatment)) +
  facet_wrap(~family)+
  geom_point(aes(y = status), alpha = 0.6, position = position_jitter(height = 0.03)) +
  geom_smooth(method = "glm", method.args = list(family = "binomial"), se = FALSE) +
  scale_color_manual(values=c("cyan3", "coral"))+
  labs(
    title = "",
    y = "Probability of Mortality",
  ) +
  theme_classic();plot2

ggsave(plot2, filename="figures/final-lab-survival/survival_families.png", width=8, height=6)
```

```{r}
# Plot
plot2<-ggplot(data, aes(x = time, y = predicted_mortality, color = family, fill = family, group=family)) +
  geom_point(aes(y = status), alpha = 0.6, position = position_jitter(height = 0.03)) +
  geom_smooth(method = "glm", method.args = list(family = "binomial"), se = FALSE) +
  labs(
    title = "",
    y = "Probability of Mortality",
  ) +
  theme_classic();plot2

ggsave(plot2, filename="figures/final-lab-survival/survival_all.png", width=6, height=4)
```

There is a spread of tolerance between families.

# LT50 Analysis

Calculate the time of 50% mortality (LT50) for each family × treatment combination from the binomial model. LT50 is solved by finding where the model's predicted probability of mortality = 0.5, using population-level predictions (random effects for round excluded via `re.form = NA`).

```{r lt50-table}
# Function to find LT50 for a given family and treatment using uniroot
# re.form = NA excludes round-level random effects (population-level estimate)
get_lt50 <- function(model, family_val, treatment_val, time_range = c(0, 300)) {
  pred_fn <- function(t) {
    newdata <- data.frame(
      time      = t,
      family    = family_val,
      treatment = treatment_val
    )
    predict(model, newdata = newdata, type = "response", re.form = NA) - 0.5
  }

  # Verify the predicted curve actually crosses 0.5 within the search range
  lo <- tryCatch(pred_fn(time_range[1]), error = function(e) NA_real_)
  hi <- tryCatch(pred_fn(time_range[2]), error = function(e) NA_real_)
  if (anyNA(c(lo, hi)) || sign(lo) == sign(hi)) return(NA_real_)

  tryCatch(
    uniroot(pred_fn, interval = time_range)$root,
    error = function(e) NA_real_
  )
}
```

```{r}
# Compute LT50 for every family × treatment combination
lt50_results <- data %>%
  distinct(family, treatment) %>%
  rowwise() %>%
  mutate(
    LT50_hrs = get_lt50(surv_model, as.character(family), as.character(treatment))
  ) %>%
  ungroup()

# Identify the non-control treatment label automatically
treated_label <- setdiff(unique(data$treatment), "Control")

# Pivot wide so control and treated LT50 sit side by side per family
lt50_table <- lt50_results %>%
  pivot_wider(names_from = treatment, values_from = LT50_hrs) %>%
  rename(
    LT50_control  = Control,
    LT50_treated  = all_of(treated_label)
  ) %>%
  mutate(
    LT50_control  = round(LT50_control,  1),
    LT50_treated  = round(LT50_treated,  1),
    # Positive = treated takes longer to reach 50% mortality (more tolerant)
    Difference_hrs = round(LT50_treated - LT50_control, 1),
    Direction = case_when(
      is.na(Difference_hrs)   ~ "—",
      Difference_hrs > 0      ~ "Treated more tolerant (+)",
      Difference_hrs < 0      ~ "Treated less tolerant (−)",
      TRUE                    ~ "No difference"
    )
  ) %>%
  arrange(family)

knitr::kable(
  lt50_table,
  caption = "LT50 (hours to 50% mortality) by family and treatment group",
  col.names = c("Family", "LT50 Control (hrs)", "LT50 Treated (hrs)",
                "Difference (hrs)", "Direction")
)
```  

Family 163: Treated died earlier (p=0.037) - LT50 is 2.8 hrs less in treated group
Family 153: Treated died later (p=0.0009) - LT50 is 6 hours more in treated group
Family 137: Treated died later (p=0.004) - LT50 is 6.3 hours more in treated group
Family 116: Treated died earlier (p=0.003) - LT50 is 5.4 hours less in treated group 
