---
title: Final lab survival analysis
author: "AS Huffmyer"
date: '2026'
output:
  html_document:
    code_folding: hide
    toc: yes
    toc_depth: 6
    toc_float: yes
  pdf_document:
    keep_tex: yes
editor_options: 
  chunk_output_type: console
---

# Set up 

Set up workspace, set options, and load required packages.    
```{r}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Load libraries. 
```{r}
library(tidyverse)
library(ggplot2)
library(survival)
library(readxl)
library(ggsurvfit)
library(gtsummary)
library(cardx)
library(cowplot)
```

# Read in data 

Read in data. 
```{r}
data<-read_csv("data/final-lab-testing/final_lab_survival.csv")%>%
  mutate(time=as.numeric(time))
```

Read in metadata. 
```{r}
metadata<-read_csv("data/final-lab-testing/bag_metadata.csv")%>%select(!c(field_tags,live, dead))
```

Merge in metadata. 
```{r}
data<-left_join(data, metadata)
head(data)
```

"treatment" here refers to the original hardening treatment. All oysters were run in the survival assay at 34-36°C.  

## Generate Kaplan Meier survival curves

```{r}
s1 <- survival::survfit(Surv(time, status) ~ treatment+family, data = data)
str(s1)
```

Plot the survival function
```{r}
ggsurvfit::survfit2(Surv(time, status) ~ treatment+family, data = data) %>% 
  ggsurvfit::ggsurvfit() +
  labs(
    x = "Hours",
    y = "Survival probability"
  )
```

Estimate survival at 24 hours. 
```{r}
summary(survival::survfit(Surv(time, status) ~ treatment+family, data = data), times = 24)
```

Use a log rank model to determine statistical differences in curves. 
```{r}
survival::survdiff(Surv(time, status) ~ treatment+family, data = data)
```

Analyze again with a Cox proportional hazards model. 
```{r}
survival::coxph(Surv(time, status) ~ treatment*family, data = data)

survival::coxph(Surv(time, status) ~ treatment*family, data = data) %>% 
  tbl_regression(exp = TRUE) 
```

## Generate plots 
 
Set theme. 
```{r}
my_theme<-theme_classic()+
  theme(
    axis.title.y = element_text(size = 14, color = "black"),
    axis.text.y = element_text(size=12, color = "black"),
    axis.text.x = element_text(size=12, color = "black"),
    axis.title.x =element_text(size = 14, color = "black"),
    legend.position = "none")
```

```{r}
plot1<-ggsurvfit::survfit2(Surv(time, status) ~ treatment, data = data) %>% 
  ggsurvfit::ggsurvfit() +
  labs(
    x = "Hours",
    y = "Survival probability",
    title=""
  )+
  ylim(0,1)+
  my_theme+
  #scale_linetype_manual(values=c("solid", "dashed"))+
  scale_colour_manual(values=c("darkgray", "orange"))+
  theme(legend.position="right");plot1
```

By family
```{r}
plot2<-ggsurvfit::survfit2(Surv(time, status) ~ family, data = data) %>% 
  ggsurvfit::ggsurvfit() +
  labs(
    x = "Hours",
    y = "Survival probability",
    title=""
  )+
  ylim(0,1)+
  my_theme+
  theme(legend.position="right");plot2
```

By group
```{r}
plot3<-ggsurvfit::survfit2(Surv(time, status) ~ family+treatment, data = data) %>% 
  ggsurvfit::ggsurvfit() +
  labs(
    x = "Hours",
    y = "Survival probability",
    title=""
  )+
  ylim(0,1)+
  my_theme+
  theme(legend.position="right");plot3
```
